# Chi-Square Cross-Tabulation

## The Worked Example

```{r, message=FALSE, warning=FALSE, echo=FALSE}

dat_o <- tibble(Groups = c("A", "A","B","B"),
                Response = c("Yes","No","Yes","No"),
                Counts = c(88,93,160,59)) %>% 
  pivot_wider(names_from = "Response", values_from = "Counts")

#md_d <- dat_o %>% 
#  pivot_longer(cols = A:D,
#               names_to = "Conditions", 
#               values_to = "Counts") %>% 
#  pull(Counts)

#colsMax <- which(md_d == max(md_d))

```

Here is our data:

```{r, echo=FALSE}
knitr::kable(dat_o, align = "c")
```

We are going to need to know the Column Totals and Row Totals and the Total number of participants (N), so lets calculate them add them to our tables:

* **Group A Row Total** = `r dat_o %>% filter(Groups == "A") %>% pull(Yes)` + `r dat_o %>% filter(Groups == "A") %>% pull(No)` = `r dat_o %>% filter(Groups == "A") %>% pull(Yes) + dat_o %>% filter(Groups == "A") %>% pull(No)`
* **Group B Row Total** = `r dat_o %>% filter(Groups == "B") %>% pull(Yes)` + `r dat_o %>% filter(Groups == "B") %>% pull(No)` = `r dat_o %>% filter(Groups == "B") %>% pull(Yes) + dat_o %>% filter(Groups == "B") %>% pull(No)`
* **Yes Column Total** = `r dat_o %>% filter(Groups == "A") %>% pull(Yes)` + `r dat_o %>% filter(Groups == "B") %>% pull(Yes)` = `r dat_o %>% filter(Groups == "A") %>% pull(Yes) + dat_o %>% filter(Groups == "B") %>% pull(Yes)`
* **No Column Total** = `r dat_o %>% filter(Groups == "A") %>% pull(No)` + `r dat_o %>% filter(Groups == "B") %>% pull(No)` = `r dat_o %>% filter(Groups == "A") %>% pull(No) + dat_o %>% filter(Groups == "B") %>% pull(No)`
* **N =** `r dat_o %>% filter(Groups == "A") %>% pull(Yes)` + `r dat_o %>% filter(Groups == "A") %>% pull(No)` +`r dat_o %>% filter(Groups == "B") %>% pull(Yes)` + `r dat_o %>% filter(Groups == "B") %>% pull(No)` = `r dat_o %>% filter(Groups == "A") %>% pull(Yes) + dat_o %>% filter(Groups == "A") %>% pull(No) + dat_o %>% filter(Groups == "B") %>% pull(Yes) + dat_o %>% filter(Groups == "B") %>% pull(No)`

And if we add those to our table we see:

```{r, echo=FALSE}

colsums_chi <- tibble(Response = c("Yes","No"),
               sum = c(dat_o %>% filter(Groups == "A") %>% pull(Yes) + dat_o %>% filter(Groups == "B") %>% pull(Yes),dat_o %>% filter(Groups == "A") %>% pull(No) + dat_o %>% filter(Groups == "B") %>% pull(No))) %>%
  pivot_wider(names_from = "Response",
              values_from = "sum") %>%
  mutate(Groups = "Totals") %>%
  select(Groups, Yes, No)

dat <- dat_o %>%
  bind_rows(colsums_chi) %>%
  mutate(Totals = Yes + No)

knitr::kable(dat, align = "c")
```

Now the formula for the chi-square is:

$$\chi^2 = \sum\frac{(Observed - Expected)^2}{Expected}$$

The Expected values for each condition, in the cross-tabulation in a few ways but the one we will use here is probably the easiest to use and it is:


$$Expected = \frac{Total_{row} \times Total_{column}}{N_{total}}$$

This is the same as other versions you might have seen such as:

$$Expected = \frac{Total_{row}}{N_{total}} \times \frac{Total_{column}}{N_{total}} \times N_{total}$$

The will both give the same result. So using the first approach we would see that the Expected values are:

**For Group A people that said Yes:**

$$Expected_{A-Yes} = \frac{`r dat %>% filter(Groups == "A") %>% pull(Totals)` \times `r dat %>% filter(Groups == "Totals") %>% pull(Yes)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = \frac{`r format(dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes), scientific = FALSE)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = `r (dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))`$$

**For Group A people that said No:**

$$Expected_{A-No} = \frac{`r dat %>% filter(Groups == "A") %>% pull(Totals)` \times `r dat %>% filter(Groups == "Totals") %>% pull(No)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = \frac{`r format(dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No), scientific = FALSE)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = `r (dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))`$$

**For Group B people that said Yes:**

$$Expected_{B-Yes} = \frac{`r dat %>% filter(Groups == "B") %>% pull(Totals)` \times `r dat %>% filter(Groups == "Totals") %>% pull(Yes)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = \frac{`r format(dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes), scientific = FALSE)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = `r (dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))`$$

**For Group B people that said No:**

$$Expected_{B-No} = \frac{`r dat %>% filter(Groups == "B") %>% pull(Totals)` \times `r dat %>% filter(Groups == "Totals") %>% pull(No)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = \frac{`r format(dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No), scientific = FALSE)`}{`r dat %>% filter(Groups == "Totals") %>% pull(Totals)`} = `r (dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))`$$

We now have our data, let's start putting it into the formula, which we said was:

$$\chi^2 = \sum\frac{(Observed - Expected)^2}{Expected}$$

Which really means:

$$\chi^2 = \frac{(Observed_{A-Yes} - Expected_{A-Yes})^2}{Expected_{A-Yes}} + \frac{(Observed_{A-No} - Expected_{A-No})^2}{Expected_{A-No}} + \\ \frac{(Observed_{B-Yes} - Expected_{B-Yes})^2}{Expected_{B-Yes}} +  \frac{(Observed_{B-No} - Expected_{B-No})^2}{Expected_{B-No}}$$

```{r hide-we, echo=FALSE}
AYO <- dat_o %>% filter(Groups == "A") %>% pull(Yes)
ANO <- dat_o %>% filter(Groups == "A") %>% pull(No)
BYO <- dat_o %>% filter(Groups == "B") %>% pull(Yes)
BNO <- dat_o %>% filter(Groups == "B") %>% pull(No)
AYE <-(dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))
ANE <-(dat %>% filter(Groups == "A") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))
BYE <-(dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(Yes))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))
BNE <-(dat %>% filter(Groups == "B") %>% pull(Totals) * dat %>% filter(Groups == "Totals") %>% pull(No))/(dat %>% filter(Groups == "Totals") %>% pull(Totals))
NT <- dat %>% filter(Groups == "Totals") %>% pull(Totals)

```

And if we start putting in the values, becomes:

$$\chi^2 = \frac{(`r AYO` - `r AYE`)^2}{`r AYE`}+ \frac{(`r ANO` - `r ANE`)^2}{`r ANE`}+\frac{(`r BYO` - `r BYE`)^2}{`r BYE`}+\frac{(`r BNO` - `r BNE`)^2}{`r BNE`}$$

And if we start to tidy those top halves up a little it becomes:

$$\chi^2 = \frac{(`r AYO - AYE`)^2}{`r AYE`}+ \frac{(`r ANO - ANE`)^2}{`r ANE`}+\frac{(`r BYO - BYE`)^2}{`r BYE`}+\frac{(`r BNO - BNE`)^2}{`r BNE`}$$

And now we square those top halves to give:

$$\chi^2 = \frac{`r (AYO - AYE)^2`}{`r AYE`} + \frac{`r (ANO - ANE)^2`}{`r ANE`} + \frac{`r (BYO - BYE)^2`}{`r BYE`} + \frac{`r (BNO - BNE)^2`}{`r BNE`} $$

And then divide the top halves by the bottom halves

$$\chi^2 = {`r ((AYO - AYE)^2)/AYE`} + {`r ((ANO - ANE)^2)/ANE`} + {`r ((BYO - BYE)^2)/BYE`} + {`r ((BNO - BNE)^2)/BNE`}$$

And then we sum them altogether to find:

$$\chi^2 = `r (((AYO - AYE)^2)/AYE) + (((ANO - ANE)^2)/ANE) + (((BYO - BYE)^2)/BYE) + (((BNO - BNE)^2)/BNE)` $$

Meaning that, rounded to two decimal places, we find $\chi^2 = `r ((((AYO - AYE)^2)/AYE) + (((ANO - ANE)^2)/ANE) + (((BYO - BYE)^2)/BYE) + (((BNO - BNE)^2)/BNE)) %>% round2(2)`$

**Degrees of Freedom**

The degrees of freedom for the cross-tabulation is calculated as:

$$df = (Rows - 1) \times (Columns - 1)$$

Which is read as **the number of Rows minus 1 times the number of Columns minus 1**. If we look at our original data again:

```{r df-orig, echo=FALSE}
chival <- ((((AYO - AYE)^2)/AYE) + (((ANO - ANE)^2)/ANE) + (((BYO - BYE)^2)/BYE) + (((BNO - BNE)^2)/BNE)) %>% round2(2)


knitr::kable(dat_o, align = "c")
```

Looking at the table we see we have `r dim(dat_o)[1]` rows and `r dim(dat_o)[2] - 1` columns of actual observed data (not looking at the titles and group names), so:

$$Rows - 1 = `r dim(dat_o)[1]` - 1 = `r dim(dat_o)[1] - 1`$$

And

$$Columns - 1 = `r dim(dat_o)[2]-1` - 1 = `r dim(dat_o)[2]-1 - 1`$$

Meaning that:

$$df = (Rows - 1) \times (Columns - 1)$$

which becomes:

$$df = (`r dim(dat_o)[1]` - 1) \times (`r dim(dat_o)[2]-1` - 1)$$

And reduces to:

$$df = (`r dim(dat_o)[1] - 1`) \times (`r dim(dat_o)[2]-1 - 1`)$$

leaving us with:

$$df = `r (dim(dat_o)[1] - 1) * ((dim(dat_o)[2]-1) - 1)`$$

so we see that $df = `r (dim(dat_o)[1] - 1) * ((dim(dat_o)[2]-1) - 1)`$

**The effect size**

One of the common effect sizes for a cross-tabulation chi-square test is Cramer's $V$ and is calculated as:

$$V = \sqrt\frac{\chi^2}{N \times \min(C-1, R-1)}$$

The key thing to note is $\min(C-1, R-1)$ which is read as **the minimum of EITHER the number of Columns (C) minus 1 OR the number of Rows (R) minus 1**; whichever of those two values is smallest. 

And the minimum of Columns minus 1 or Rows minus 1 is:

$$\min(C-1, R-1) = \min( `r dim(dat_o)[1]` - 1,  `r dim(dat_o)[2]-1` - 1)$$
$$\min(C-1, R-1) = \min( `r dim(dat_o)[1] - 1`,  `r dim(dat_o)[2]-1 - 1`)$$
which gives us:

$$\min(C-1, R-1) = `r min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))`$$

And now we can start completing the Cramer's $V$ formula as we know:

* $\chi^2 = `r chival`$
* $N = `r NT`$
* $\min(C-1, R-1) = `r min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))`$

Giving us:

$$V = \sqrt\frac{`r chival`}{`r NT` \times `r min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))`}$$

And if we deal with the bottom half first we get:

$$V = \sqrt\frac{`r chival`}{`r NT * (min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1)))`}$$

Then if we divide the top by the bottom we get

$$V = \sqrt{`r chival /(NT * (min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))))`}$$

Giving us:

$$V = `r sqrt(chival /(NT * (min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1)))))`$$

So we see that, rounded to two decimal places, the effect size is $V = `r sqrt(chival /(NT * (min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))))) %>% round2(2)`$

**The write-up**

```{r writeup, echo=FALSE}

df <- (dim(dat_o)[1] - 1) * ((dim(dat_o)[2]-1) - 1)
V <- sqrt(chival /(NT * (min((dim(dat_o)[1] - 1), (dim(dat_o)[2]-1 - 1))))) %>% round2(2)
```

If we were to look at a critical value look-up table, we would see that the critical value associated with a $df = `r df`$ at $\alpha = .05$, to three decimal places, is $\chi^2_{crit} = `r qchisq(.05, df, lower.tail = FALSE) %>% round2(3)`$. As the chi-square value of this test (i.e. $\chi^2 = `r chival`$) is `r ifelse(chival >= (qchisq(.05, df, lower.tail = FALSE) %>% round2(3)),"larger","smaller")` than $\chi^2_{crit}$ then we can say that our test is `r ifelse(chival >= (qchisq(.05, df, lower.tail = FALSE) %>% round2(3)),"significant","non-significant")`, and as such would be written up as $\chi^2(df = `r df`, N = `r NT`) = `r chival`,`r ifelse(chival >= (qchisq(.05, df, lower.tail = FALSE) %>% round2(3)),"p < .05","p > .05")`, V = `r V`$.

If our test was significant then we would go on to carry out the relevant one-sample chi-squares to breakdown how the association manifests itself. The key thing to remember however is that in these one-sample chi-squares, following the cross-tabulation, you **MUST** use the expected values from the cross-tabulation and **do not** calculate new expected values. For instance if you were looking at the one-sample chi-square within Group A, then you would use the expected values of $Expected_{A-Yes} = `r AYE`$ and $Expected_{A-No} = `r ANE`$ to compare against the observed values of $Observed_{A-Yes} = `r AYO`$ and $Observed_{A-No} = `r ANO`$ as such:

$$\chi^2 = \frac{(`r AYO` - `r AYE`)^2}{`r AYE`}+ \frac{(`r ANO` - `r ANE`)^2}{`r ANE`}$$

Which if you walk the process through you find:

$$\chi^2 = `r (((AYO - AYE)^2)/AYE) + (((ANO - ANE)^2)/ANE)`$$

Meaning that, rounded to two decimal places, we find $\chi^2 = `r ((((AYO - AYE)^2)/AYE) + (((ANO - ANE)^2)/ANE)) %>% round2(2)`$


**To Be Continued**

## Test Yourself

### DataSet 1

To Be Added

## ChiSquare Look-up Table


|df | $\alpha = .05$|
|:-:|:-------------:|
|1  |`r qchisq(.05, 1, lower.tail = FALSE) %>% round2(3)`|
|2  |`r qchisq(.05, 2, lower.tail = FALSE) %>% round2(3)`|
|3  |`r qchisq(.05, 3, lower.tail = FALSE) %>% round2(3)`|
|4  |`r qchisq(.05, 4, lower.tail = FALSE) %>% round2(3)`|
|5  |`r qchisq(.05, 5, lower.tail = FALSE) %>% round2(3)`|
|6  |`r qchisq(.05, 6, lower.tail = FALSE) %>% round2(3)`|
|7  |`r qchisq(.05, 7, lower.tail = FALSE) %>% round2(3)`|
|8  |`r qchisq(.05, 8, lower.tail = FALSE) %>% round2(3)`|
|9  |`r qchisq(.05, 9, lower.tail = FALSE) %>% round2(3)`|
|10  |`r qchisq(.05, 10, lower.tail = FALSE) %>% round2(3)`|
