# Descriptives

## A note on Descriptives and Inferentials

In the first part of this book we are going to look at `r glossary("descriptive", display = "descriptive statistics")` which we use to summarise and describe data we have collected - normally from a `r glossary("sample")` of the `r glossary("population")` of interest. Remember that the population is all the members of a group that we wish to generalise our findings to, and the sample is the subset of that population we have gathered data from. We draw our sample from the population, so to speak. The population is what we determine it to be based on our research question - everyone in Psychology at Glasgow; everyone study Psychology in Scotland; everyone studying Psychology in the world! Most of the time, testing everyone in a population is an unrealistic goal so we take a sample of that population and make inferences from that sample to the population using statistics. And the first step we take is to describe our sample.

Descriptive statistics can be broken down into three kinds:

* `r glossary::glossary("point estimates", def = "a single value that summarises an aspect of a data set.  Examples include the mean, median, and the mode.", link = FALSE)` - these tend to be single values, or points, that summarise a set of data. Most common would be ones that summarise the center of the data such as the mean and the median.
* `r glossary::glossary("interval estimates", def = "a range of values that summarise an aspect of a data set.  Examples include the range, variance, standard deviation, standard error and confidence intervals.", link = FALSE)` - these tend to be a range of values that summarise a set of data. Most common would be ones that summarise the spread of the data such as the variance and the standard deviation.
* visualisations - are figures that help display the point and interval estimates of the data and can take various different forms depending on the data type.

Once we have described the data in our sample we then use `r glossary("inferential", display = "inferential statistics")` to make predictions about, or comparisons between, our data. We infer how the general population would perform based on the sample we have collected and described. We will look at inferential statistics later in this book but first we will start with descriptives statistics, looking at point estimates first, before moving on to interval estimates.

**Point Estimates**

We will start by looking at three point estimates. They are:

* the Mean
* the Median
* the Mode

## The Mean

The `r glossary("mean")` is a descriptive statistic that measures the average value of a set of numbers.

The symbol for the mean is generally written as $\overline{X}$ (pronounced as X-bar) and the formula for the mean is:

$$\overline{X} = \frac{\sum_i^n{x_i}}{n}$$

Which reads as sum ($\sum$) all the values from $i$ (the first value) to $n$ (the last value) and then divide that number by the total number of values ($n$). Let's look at an example.

Say we are interested in how years of experience in driving impacts on some metric of peformance. We recruit 25 `r glossary::glossary("participant", def = "the word used to describe someone who has taken part in a study. Note that subject is outdated and no longer used.", display = "participants", link = FALSE, add_to_table = TRUE)` and ask them how many years they have been driving. Here are their responses:

```{r driving-years, echo = FALSE}
drv <- tibble(Participant = 1:25,
              Years = c(7, 1, 2, 6, 3, 4, 3, 4, 3, 4, 5, 4, 7, 5, 6, 5, 5, 4, 5, 6, 5, 6, 3, 2, 5
))
```

$$`r drv %>% pull(Years)`$$

If we start to fill in the information from above we get:

$$\overline{X} = \frac{`r drv %>% filter(Participant == 1) %>% pull(Years)` +
`r drv %>% filter(Participant == 2) %>% pull(Years)` +
`r drv %>% filter(Participant == 3) %>% pull(Years)`+
`r drv %>% filter(Participant == 4) %>% pull(Years)`+
`r drv %>% filter(Participant == 5) %>% pull(Years)`+\\
`r drv %>% filter(Participant == 6) %>% pull(Years)`+
`r drv %>% filter(Participant == 7) %>% pull(Years)`+
`r drv %>% filter(Participant == 8) %>% pull(Years)`+
`r drv %>% filter(Participant == 9) %>% pull(Years)`+
`r drv %>% filter(Participant == 10) %>% pull(Years)`+\\
`r drv %>% filter(Participant == 11) %>% pull(Years)`+
`r drv %>% filter(Participant == 12) %>% pull(Years)`+
`r drv %>% filter(Participant == 13) %>% pull(Years)`+
`r drv %>% filter(Participant == 14) %>% pull(Years)`+
`r drv %>% filter(Participant == 15) %>% pull(Years)`+\\
`r drv %>% filter(Participant == 16) %>% pull(Years)`+
`r drv %>% filter(Participant == 17) %>% pull(Years)`+
`r drv %>% filter(Participant == 18) %>% pull(Years)`+
`r drv %>% filter(Participant == 19) %>% pull(Years)`+
`r drv %>% filter(Participant == 20) %>% pull(Years)`+\\
`r drv %>% filter(Participant == 21) %>% pull(Years)`+
`r drv %>% filter(Participant == 22) %>% pull(Years)`+
`r drv %>% filter(Participant == 23) %>% pull(Years)`+
`r drv %>% filter(Participant == 24) %>% pull(Years)`+
`r drv %>% filter(Participant == 25) %>% pull(Years)`}{`r dim(drv)[1]`}$$

So the first step is to add all the values on the top half of the formula together, which is called the `r glossary::glossary("numerator", def = "name for the top half of a formula", link = FALSE)`, giving us:

$$\overline{X} =\frac{`r drv %>% pull(Years) %>% sum()`}{`r dim(drv)[1]`}$$

Which can also be read as:

$$\overline{X} = `r drv %>% pull(Years) %>% sum()`/`r dim(drv)[1]`$$

And if we then divide top of the formula, the numerator, by the bottom half, called the `r glossary::glossary("denominator", def = "name for the bottom half of a formula", link = FALSE)`, we would get:

$$\overline{X} = `r (drv %>% pull(Years) %>% sum())/(dim(drv)[1])`$$

And so, when using the standard APA write-up format of **M = ...**, we would write **M = `r (drv %>% pull(Years) %>% sum())/(dim(drv)[1])`**.

### Test Yourself - The Mean

Here are 5 sets of data to practice calculating the mean on. Calculate the mean on each one and then answer the questions below to see if you were correct.

```{r t1, echo=FALSE}
set.seed(1409)
dat1 <- sample(1:30, 10, replace = TRUE)
dat2 <- sample(1:30, 10, replace = TRUE)
dat3 <- sample(1:30, 10, replace = TRUE)
dat4 <- sample(1:30, 10, replace = TRUE)
dat5 <- sample(1:30, 10, replace = TRUE)

m1 <- mean(dat1)
m2 <- mean(dat2)
m3 <- mean(dat3)
m4 <- mean(dat4)
m5 <- mean(dat5)

r1 <- sample(seq(-1,1,.1),3, replace = FALSE)
r2 <- sample(seq(-1,1,.1),3, replace = FALSE)
r3 <- sample(seq(-1,1,.1),3, replace = FALSE)
r4 <- sample(seq(-1,1,.1),3, replace = FALSE)
r5 <- sample(seq(-1,1,.1),3, replace = FALSE)

options1 <- sample(c(answer = m1, m1+r1[1], m1+r1[2], m1+r1[3]))
options2 <- sample(c(answer = m2, m2+r2[1], m2+r2[2], m2+r2[3]))
options3 <- sample(c(answer = m3, m3+r3[1], m3+r3[2], m3+r3[3]))
options4 <- sample(c(answer = m4, m4+r4[1], m4+r4[2], m4+r4[3]))
options5 <- sample(c(answer = m5, m5+r5[1], m5+r5[2], m5+r5[3]))

```

* DataSet 1 - $`r dat1`$
* DataSet 2 - $`r dat2`$
* DataSet 3 - $`r dat3`$
* DataSet 4 - $`r dat4`$
* DataSet 5 - $`r dat5`$
<p></p>
1. What is the mean of DataSet 1? `r mcq(options1)`
2. What is the mean of DataSet 2? `r mcq(options2)`
3. What is the mean of DataSet 3? `r mcq(options3)`
4. What is the mean of DataSet 4? `r mcq(options4)`
5. What is the mean of DataSet 5? `r mcq(options5)`

## The Median

The `r glossary("median")` is the next point estimate we will look at and is the middle number in a distribution where half of the values in the data are larger and half are smaller. It is literally the value that divides the data in half.

The standard way of calculating the Median involves two steps:

1. Sort all values from lowest to highest (i.e. from $i$ (the first value) to $n$ (the last value))
2. The median is the value at position $\frac{(n + 1)}{2}$

So if we look at our driving data again, the responses were:

$$`r drv %>% pull(Years)`$$

And if we sort them from lowest value to highest value we get:

$$`r drv %>% arrange(Years) %>% pull(Years)`$$

Now we need to figure out what the value would be at the median position of our dataset, where the median position is $\frac{(n + 1)}{2}$ and we have n = `r dim(drv)[1]` participants.

$$Median\space Position = \frac{(n + 1)}{2} = \frac{(`r dim(drv)[1]` + 1)}{2} = \frac{`r dim(drv)[1] + 1`}{2} = `r (dim(drv)[1] + 1)/2`$$

Meaning that the Median Position for n = `r dim(drv)[1]` is the `r (dim(drv)[1] + 1)/2`^th^ position and so the Median is the value at position `r (dim(drv)[1] + 1)/2` after we have sorted the data from smallest to largest:

$$`r drv %>% arrange(Years) %>% pull(Years)`$$

And if we count along the line until we get to the Median Position, we can see that the value in position `r (dim(drv)[1] + 1)/2` is `r drv %>% arrange(Years) %>% slice(((dim(drv)[1] + 1)/2)) %>% pull(Years)`, meaning that for this data the Median = `r drv %>% arrange(Years) %>% slice(((dim(drv)[1] + 1)/2)) %>% pull(Years)`. And so, when using the standard APA write-up format of **Mdn = ...**, we would write **Mdn = `r drv %>% arrange(Years) %>% slice(((dim(drv)[1] + 1)/2)) %>% pull(Years)`**.

### Test Yourself - The Median

Here are 5 sets of data to practice calculating the median on. Calculate the median on each one and then answer the questions below to see if you were correct.


```{r median1, echo = FALSE}
set.seed(1409)

dat2 <- sample(rep(c(1:5), times = sample(c(1:5), replace = FALSE)), replace = FALSE)
dat1 <- sample(rep(c(6:10), times = sample(c(2,4,3,3,1), replace = FALSE)), replace = FALSE)
dat3 <- sample(rep(sample(rep(seq(0,50,1), each = 1),5, replace = TRUE), times = c(3,3,2,1,6), replace = FALSE), replace = FALSE)
dat5 <- sample(rep(sample(rep(seq(10,20), each = 1),5, replace = TRUE), times = c(1,2,3,2,4), replace = FALSE), replace = FALSE)
dat4 <- sample(rep(sample(rep(seq(1,10,1), each = 1),5, replace = TRUE), times = c(2,2,2,2,3), replace = FALSE), replace = FALSE)


median1 <- median(dat1)
median2 <- median(dat2)
median3 <- median(dat3)
median4 <- median(dat4)
median5 <- median(dat5)

wrong1 <- sample(c(1,-1,.5,-.5),3, replace = FALSE)
wrong2 <- sample(c(1,-1,.5,-.5),3, replace = FALSE)
wrong3 <- sample(c(1,-1,.5,-.5),3, replace = FALSE)
wrong4 <- sample(c(1,-1,.5,-.5),3, replace = FALSE)
wrong5 <- sample(c(1,-1,.5,-.5),3, replace = FALSE)

options1 <- sample(c(answer = median1, median1 + wrong1[1], median1 + wrong1[2], median1 + wrong1[3]))
options2 <- sample(c(answer = median2, median2 + wrong2[1], median2 + wrong2[2], median2 + wrong2[3]))
options3 <- sample(c(answer = median3, median3 + wrong3[1], median3 + wrong3[2], median3 + wrong3[3]))
options4 <- sample(c(answer = median4, median4 + wrong4[1], median4 + wrong4[2], median4 + wrong4[3]))
options5 <- sample(c(answer = median5, median5 + wrong5[1], median5 + wrong5[2], median5 + wrong5[3]))

```

* DataSet 1 - $`r dat1`$
* DataSet 2 - $`r dat2`$
* DataSet 3 - $`r dat3`$
* DataSet 4 - $`r dat4`$
* DataSet 5 - $`r dat5`$
<p></p>
1. What is the median of DataSet 1? `r mcq(options1)`
2. What is the median of DataSet 2? `r mcq(options2)`
3. What is the median of DataSet 3? `r mcq(options3)`
4. What is the median of DataSet 4? `r mcq(options4)`
5. What is the median of DataSet 5? `r mcq(options5)`


## The Mode

The mode is the last point estimate we will consider here and is the value or category that appears most often, most frequently, in your data set. There is no formula for the mode and it involves counting how many of each category or value you have and finding the most common one.  Here are our values again:

$$`r drv %>% pull(Years)`$$

And if we sort them from largest to smallest for easy of reading we get:

$$`r drv %>% arrange(Years) %>% pull(Years)`$$

And if we start to count all the different values we see that we have:

```{r mode-tab, echo=FALSE}
modtab <- drv %>% group_by(Years) %>% count()

threc <- modtab %>% filter(Years == 3) %>% pull(n)
fourc <- modtab %>% filter(Years == 4) %>% pull(n)

maxc <- modtab %>% arrange(desc(n)) %>% ungroup %>% slice(1) %>% pull(Years)
pepc <- modtab %>% arrange(desc(n)) %>% ungroup %>% slice(1) %>% pull(n)
  
knitr::kable(modtab)
```

Meaning, for example, we have `r threc` people with three years of driving experience, and `r fourc` people with four years of driving experience. And looking at our table we can see the most common number of years of driving experience is `r maxc` as there are `r pepc` with that number of years experience. We would therefore write **Mode = `r maxc`**

### Test Yourself - The Mode

Here are 5 sets of data to practice calculating the mode on. Calculate the mean on each one and then answer the questions below to see if you were correct.

```{r mode1, echo = FALSE}
set.seed(1409)

dat1 <- sample(rep(c(1:5), times = sample(c(1:5), replace = FALSE)), replace = FALSE)
dat2 <- sample(rep(c(6:10), times = sample(c(2,4,3,3,1), replace = FALSE)), replace = FALSE)
dat3 <- sample(rep(sample(rep(seq(0,50,1), each = 1),5, replace = TRUE), times = c(3,3,2,1,6), replace = FALSE), replace = FALSE)
dat4 <- sample(rep(sample(rep(seq(10,20), each = 1),5, replace = TRUE), times = c(1,2,3,2,4), replace = FALSE), replace = FALSE)
dat5 <- sample(rep(sample(rep(seq(1,10,1), each = 1),5, replace = TRUE), times = c(2,2,2,2,3), replace = FALSE), replace = FALSE)


mode1 <- dat1 %>% tibble(x = .) %>% group_by(x) %>% count() %>% ungroup() %>% filter(n == max(n)) %>% pull(x)
mode2 <- dat2 %>% tibble(x = .) %>% group_by(x) %>% count() %>% ungroup() %>% filter(n == max(n)) %>% pull(x)
mode3 <- dat3 %>% tibble(x = .) %>% group_by(x) %>% count() %>% ungroup() %>% filter(n == max(n)) %>% pull(x)
mode4 <- dat4 %>% tibble(x = .) %>% group_by(x) %>% count() %>% ungroup() %>% filter(n == max(n)) %>% pull(x)
mode5 <- dat5 %>% tibble(x = .) %>% group_by(x) %>% count() %>% ungroup() %>% filter(n == max(n)) %>% pull(x)

wrong1 <- tibble(y = unique(dat1)) %>% filter(y != mode1) %>% pull(y)
wrong2 <- tibble(y = unique(dat2)) %>% filter(y != mode2) %>% pull(y)
wrong3 <- tibble(y = unique(dat3)) %>% filter(y != mode3) %>% pull(y)
wrong4 <- tibble(y = unique(dat4)) %>% filter(y != mode4) %>% pull(y)
wrong5 <- tibble(y = unique(dat5)) %>% filter(y != mode5) %>% pull(y)

options1 <- sample(c(answer = mode1, wrong1[1], wrong1[2], wrong1[3]))
options2 <- sample(c(answer = mode2, wrong2[1], wrong2[2], wrong2[3]))
options3 <- sample(c(answer = mode3, wrong3[1], wrong3[2], wrong3[3]))
options4 <- sample(c(answer = mode4, wrong4[1], wrong4[2], wrong4[3]))
options5 <- sample(c(answer = mode5, wrong5[1], wrong5[2], wrong5[3]))

```

* DataSet 1 - $`r dat1`$
* DataSet 2 - $`r dat2`$
* DataSet 3 - $`r dat3`$
* DataSet 4 - $`r dat4`$
* DataSet 5 - $`r dat5`$
<p></p>
1. What is the mode of DataSet 1? `r mcq(options1)`
2. What is the mode of DataSet 2? `r mcq(options2)`
3. What is the mode of DataSet 3? `r mcq(options3)`
4. What is the mode of DataSet 4? `r mcq(options4)`
5. What is the mode of DataSet 5? `r mcq(options5)`


## A full desciptive example

**The Data**

To look at a full descriptive example including the mean, variance, standard deviation, standard error, 95% Confidence Intervals, and z-scores, we will use the following data from 10 participants:

```{r we, message=FALSE, warning=FALSE, echo=FALSE}

dat_o <- tibble(Participants = 1:10,
              PreTest = c(60, 64, 56, 82, 74, 79, 63, 59, 72, 66),
              PostTest = c(68, 75, 62, 85, 73, 85, 64, 59, 73, 70))

dat <- dat_o %>% mutate(X = abs(PostTest - PreTest)) %>% select(Participants, X)

meanx <- dat %>% pull(X) %>% mean()
varx <- dat %>% pull(X) %>% var()
sdx <- dat %>% pull(X) %>% sd()
nx <- dim(dat)[1]
semx <- sdx/sqrt(nx)
zci <- 1.96

knitr::kable(dat, align = "c")
```

### The Mean

The symbol for the mean is generally written as $\overline{X}$ (pronounced as X-bar) and the formula for the mean is:

$$\overline{X} = \frac{\sum_i^n{x_i}}{n}$$

Which reads as sum ($\sum$) all the values from $i$ (the first value) to $n$ (the last value) and then divide that number by the total number of values ($n$). If we start to fill in the information from above we get:

$$\overline{X} = \frac{`r dat %>% filter(Participants == 1) %>% pull(X)` + `r dat %>% filter(Participants == 2) %>% pull(X)` + `r dat %>% filter(Participants == 3) %>% pull(X)` +`r dat %>% filter(Participants == 4) %>% pull(X)` +`r dat %>% filter(Participants == 5) %>% pull(X)` +`r dat %>% filter(Participants == 6) %>% pull(X)` +`r dat %>% filter(Participants == 7) %>% pull(X)` +`r dat %>% filter(Participants == 8) %>% pull(X)` +`r dat %>% filter(Participants == 9) %>% pull(X)` +`r dat %>% filter(Participants == 10) %>% pull(X)`}{`r dim(dat)[1]`}$$

And if we sum all the values on the top half together we get:

$$\overline{X} = \frac{`r dat %>% pull(X) %>% sum()`}{`r dim(dat)[1]`}$$

And finally divide the top half by the bottom half, leaving:

$$\overline{X} = `r (dat %>% pull(X) %>% sum())/(dim(dat)[1])`$$

We find that the mean, rounded to two decimal places, is **$\overline{X} = `r meanx`$**

### Variance

The symbol for the variance is generally written as $s^2$ (pronounced as sigma-squared) but can also be written as $Var$. The formula for the variance is:

$$s^2 = \frac{\sum_i^n(x_{i} - \overline{x})^2}{n-1}$$

So let's again start by filling in the numbers to the formula:

$$s^2 = \frac{(`r dat %>% filter(Participants == 1) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 2) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 3) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 4) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 5) %>% pull(X)` - `r meanx`)^2 + \\(`r dat %>% filter(Participants == 6) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 7) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 8) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 9) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 10) %>% pull(X)` - `r meanx`)^2}{`r dim(dat)[1]` - 1}$$

And we do the subtractions in all those brackets of the top half and the one on the bottom:

$$s^2 = \frac{(`r dat %>% filter(Participants == 1) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 2) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 3) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 4) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 5) %>% pull(X) - meanx`)^2 + \\(`r dat %>% filter(Participants == 6) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 7) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 8) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 9) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 10) %>% pull(X) - meanx`)^2}{`r dim(dat)[1]-1`}$$

Then we square all those values on the top half:

$$s^2 = \frac{`r (dat %>% filter(Participants == 1) %>% pull(X) - meanx)^2` + `r (dat %>% filter(Participants == 2) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 3) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 4) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 5) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 6) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 7) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 8) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 9) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 10) %>% pull(X) - meanx)^2`}{`r dim(dat)[1]-1`}$$

Sum those values together:

$$s^2 = \frac{`r ((dat %>% filter(Participants == 1) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 2) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 3) %>% pull(X) - meanx)^2)  + ((dat %>% filter(Participants == 4) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 5) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 6) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 7) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 8) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 9) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 10) %>% pull(X) - meanx)^2)`}{`r dim(dat)[1]-1`}$$
And divide the top half by the bottom half:

$$s^2 = `r dat %>% pull(X) %>% var()`$$

Showing that the variance, rounded to two decimal places is, **$s^2 = `r dat %>% pull(X) %>% var() %>% round2(2)`$**

### Standard Deviation

The symbol for the standard deviation is generally written as $s$ (pronounced as sigma) but can also be written as $SD$. The formula for the standard deviation is:

$$s = \sqrt\frac{\sum_i^n(x_{i} - \overline{x})^2}{n-1}$$
So let's again start by filling in the numbers to the formula:

$$s = \sqrt\frac{(`r dat %>% filter(Participants == 1) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 2) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 3) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 4) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 5) %>% pull(X)` - `r meanx`)^2 + \\(`r dat %>% filter(Participants == 6) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 7) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 8) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 9) %>% pull(X)` - `r meanx`)^2 + (`r dat %>% filter(Participants == 10) %>% pull(X)` - `r meanx`)^2}{`r dim(dat)[1]` - 1}$$

And we do the subtractions in all those brackets of the top half and the one on the bottom:

$$s = \sqrt\frac{(`r dat %>% filter(Participants == 1) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 2) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 3) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 4) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 5) %>% pull(X) - meanx`)^2 + \\(`r dat %>% filter(Participants == 6) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 7) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 8) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 9) %>% pull(X) - meanx`)^2 + (`r dat %>% filter(Participants == 10) %>% pull(X) - meanx`)^2}{`r dim(dat)[1]-1`}$$

Then we square all those values on the top half:

$$s = \sqrt\frac{`r (dat %>% filter(Participants == 1) %>% pull(X) - meanx)^2` + `r (dat %>% filter(Participants == 2) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 3) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 4) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 5) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 6) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 7) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 8) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 9) %>% pull(X) - meanx)^2`+ `r (dat %>% filter(Participants == 10) %>% pull(X) - meanx)^2`}{`r dim(dat)[1]-1`}$$

Sum those values together:

$$s = \sqrt\frac{`r ((dat %>% filter(Participants == 1) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 2) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 3) %>% pull(X) - meanx)^2)  + ((dat %>% filter(Participants == 4) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 5) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 6) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 7) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 8) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 9) %>% pull(X) - meanx)^2) + ((dat %>% filter(Participants == 10) %>% pull(X) - meanx)^2)`}{`r dim(dat)[1]-1`}$$
Divide the top half by the bottom half:

$$s = \sqrt{`r varx`}$$

And finally take the square root:

$$s = `r sdx`$$

We find that the standard deviation, rounded to two decimal places, is **$s = `r sdx %>% round2(2)`$**

### Var to SD

Remembering that the standard deviation is the square root of the variance, then if you know the variance ($s^2$) and you need the standard deviation ($s$) then you can do:

$$s = \sqrt{s^2}$$

In our example, $s^2 = `r varx`$ and so

$$s = \sqrt{`r varx`}$$

giving:

$$s = `r sqrt(varx)`$$

### SD to Var

Sometimes you need to go from the standard deviation to the variance, and remembering that the variance is the standard deviation squared, or the standard deviation multiplied by itself, then you can do:

$$s^2 = s \times s$$

And if we know that the standard deviation in our example is $s = `r sdx`$ then:

$$s^2 = `r sdx` \times `r sdx`$$

giving us:

$$s^2 = `r sdx * sdx`$$

### Z-Scores

Any value on a continuous scale can be converted to a z-score (standard deviation units) through the formula:

$$z = \frac{x - \overline{x}}{s_{x}}$$

which can be read as the value ($x$) minus the mean value of the data ($\overline{x}$), divided by the standard deviation of the data ($s_{X}$).  For example, if we look at Participant 1:

* $x = `r dat %>% filter(Participants == 1) %>% pull(X)`$
* $\overline{x} = `r meanx`$
* $s_{x} = `r sdx`$

Which if we fill in the formula we get:

$$z = \frac{`r dat %>% filter(Participants == 1) %>% pull(X)` - `r meanx`}{`r sdx`}$$

And if we sort our the top half first:

$$z = \frac{`r dat %>% filter(Participants == 1) %>% pull(X) - meanx`}{`r sdx`}$$

Leaving us with:

$$z = `r (dat %>% filter(Participants == 1) %>% pull(X) - meanx)/sdx`$$

So Participant 1, when converted to a z-score and rounded to two decimal places, would have **z = `r ((dat %>% filter(Participants == 1) %>% pull(X) - meanx)/sdx) %>% round2(2)`**

### Standard Error of the Mean

The symbol for the standard error is usually $SE$ but can also be written as $SEM$ when specificaly about the mean - short for standard error of the mean.  The formula for the standard error is the standard deviation ($s$) divided by the square root of the number of observations ($\sqrt{n}$), and is written as:

$$SE = \frac{s}{\sqrt{n}}$$

but can also be written as:

$$SE = \frac{SD}{\sqrt{n}}$$

We know that our $SD = `r sdx`$ and that we have $n = `r nx`$ participants, so:

$$SE = \frac{`r sdx`}{\sqrt{`r nx`}}$$

And if we do the square root of the bottom half (denominator):

$$SE = \frac{`r sdx`}{`r sqrt(nx)`}$$

And divide the top by the bottom:

$$SE = `r semx`$$

We find that the standard error, rounded to two decimal places, is $SE = `r semx %>% round2(2)`$

### Confidence Intervals

We will specifically focus on the 95% Confidence Interval using the cut-off value (assuming $\alpha = .05$ and two-tailed) of $z = 1.96$. The key formulas are:

$$Upper \space 95\%CI = \overline{x} + (z \times SE)$$

And

$$Lower \space 95\%CI = \overline{X} - (z \times SE)$$

We know that:

* $\overline{x} = `r meanx`$
* $SE = `r semx`$
* $z = `r zci`$

**Upper 95% CI**

Dealing with the **Upper 95% CI** we get:

$$Upper \space 95\%CI = `r meanx` + (`r zci` \times `r semx`)$$

And if we sort out the bracket first:

$$Upper \space 95\%CI = `r meanx` + `r zci * semx`$$

Which leaves us with:

$$Upper \space 95\%CI = `r meanx + (zci * semx)`$$

**Lower 95% CI**

And now the **Lower 95% CI** we get:

$$Lower \space 95\%CI = `r meanx` - (`r zci` \times `r semx`)$$

And if we sort out the bracket first:

$$Lower \space 95\%CI = `r meanx` - `r zci * semx`$$

Which leaves us with:

$$Lower \space 95\%CI = `r meanx - (zci * semx)`$$

And if we round both those values to two decimal places, then you get **$Lower \space 95\%CI = `r (meanx - (zci * semx)) %>% round2(2)`$** and **$Upper \space 95\%CI = `r (meanx + (zci * semx)) %>% round2(2)`$**

### The Write-Up

And if we were to do a decent presentation of this data, including a measure of central tendency (the mean), a measure of spread relating to the sample (the standard deviation), and a measure of spread relating to the population (the 95% CI), then it would start as something like: **"We ran `r nx` participants (M = `r meanx %>% round2(2)`, SD = `r sdx %>% round2(2)`, 95%CI = [`r (meanx - (zci * semx)) %>% round2(2)`, `r (meanx + (zci * semx)) %>% round2(2)`])....."**

## Section glossary

```{r gloss-tab, echo=FALSE, results='asis'}

glossary_table()
```

