# Between-Subjects Welch's t-test

**between-subjects Welch's t-test:** Compare two groups or conditions where the participants are different in each group and have not been matched or are only matched on broad demographics, e.g. only age.

## The Worked Example

```{r we, echo = FALSE}
set.seed(1410)

mx <- runif(1, 0, 100) %>% round2(2)
sdx <- runif(1, 0, 5) %>% round2(2)
N <- seq(10, 100, 2) %>% sample(1)

dat <- tibble(Group = rep(c("A","B"), each = N/2),
              scores = rnorm(N, mx, sdx)) %>%
  mutate(scores = scores + rep(c(0,2), each = N/2))

desc <- dat %>% 
  group_by(Group) %>%
  summarise(Npg = n(),
            Means = mean(scores) %>% round2(2),
            SDs = sd(scores) %>% round2(2),
            Variance = var(scores) %>% round2(2),
            `Standard Err.` = (SDs/sqrt(Npg)) %>% round2(2))

m_a <- desc %>% filter(Group == "A") %>% pull(Means)
m_b <- desc %>% filter(Group == "B") %>% pull(Means)
sd_a <- desc %>% filter(Group == "A") %>% pull(SDs)
sd_b <- desc %>% filter(Group == "B") %>% pull(SDs)
var_a <- sd_a^2
var_b <- sd_b^2
n_a <- desc %>% filter(Group == "A") %>% pull(Npg)
n_b <- desc %>% filter(Group == "B") %>% pull(Npg)

#cohensd <- ((2* t.value)/sqrt(N-2)) %>% round2(2)

results <- t.test(dat %>% filter(Group == "A") %>% pull(scores), 
                  dat %>% filter(Group == "B") %>% pull(scores), 
                  var.equal = FALSE) %>% 
  tidy() %>%
  mutate(cohensd_t = (2*statistic)/sqrt(parameter))
```

Here is your data:

```{r, echo = FALSE}
desc %>% 
  select(-Variance, -`Standard Err.`) %>% 
  rename(N = Npg, Mean = Means, SD = SDs) %>%
  knitr::kable(align = "c")
```

Let's look at the main t-test formula for Welch's t-test:

$$t = \frac{\bar{X}_1 - \bar{X}_2}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}$$

Now, from the table above we know:

* the mean of Group A is $\bar{X_1} = `r desc %>% filter(Group == "A") %>% pull(Means)`$, 
* the mean of Group B is $\bar{X_2} = `r desc %>% filter(Group == "B") %>% pull(Means)`$, 
* the N of Group A is $N_1 = `r desc %>% filter(Group == "A") %>% pull(Npg)`$, 
* and the N of Group B is $N_2 = `r desc %>% filter(Group == "B") %>% pull(Npg)`$

From the table above we also know the standard deviation ($s_1$ = `r desc %>% filter(Group == "A") %>% pull(SDs)` and $s_2$ = `r desc %>% filter(Group == "B") %>% pull(SDs)`) of the two groups, but the Welch's t-test formula uses the variance of both groups ($s_1^2$ and $s_2^2$). Fortunately, if you know the standard deviation of a set of values, you can calculate the variance by the formula:

$$variance = SD \times SD$$

Meaning that:

* the variance of Group A, $s_1^2 = `r desc %>% filter(Group == "A") %>% pull(SDs)` \times `r desc %>% filter(Group == "A") %>% pull(SDs)`$ = `r desc %>% filter(Group == "A") %>% pull(SDs) *  desc %>% filter(Group == "A") %>% pull(SDs)`
* and the variance of Group B, $s_2^2 = `r desc %>% filter(Group == "B") %>% pull(SDs)` \times `r desc %>% filter(Group == "B") %>% pull(SDs)`$ = `r desc %>% filter(Group == "B") %>% pull(SDs) *  desc %>% filter(Group == "B") %>% pull(SDs)`

And we can now start to fill in the t-test formula as follows:

$$t = \frac{`r m_a` - `r m_b`}{\sqrt{\frac{`r var_a`}{`r n_a`}+\frac{`r var_b`}{`r n_b`}}}$$

$$t = \frac{`r m_a` - `r m_b`}{\sqrt{`r var_a/n_a`+`r var_b/n_b`}}$$
$$t = \frac{`r m_a` - `r m_b`}{\sqrt{`r (var_a/n_a) + (var_b/n_b)`}}$$

$$t = \frac{`r m_a` - `r m_b`}{`r sqrt((var_a/n_a) + (var_b/n_b))`}$$

$$t = \frac{`r m_a - m_b`}{`r sqrt((var_a/n_a) + (var_b/n_b))`}$$

$$t = `r (m_a - m_b)/(sqrt((var_a/n_a) + (var_b/n_b)))`$$


**Degrees of Freedom**

$$df = \frac{\left(\frac{s^2_{x_1}}{n_1}+\frac{s^2_{x_2}}{n_2}\right)^2}{\frac{\left(\frac{s^2_{x_1}}{n_1}\right)^2}{n_1 - 1}+\frac{\left(\frac{s^2_{x_2}}{n_2}\right)^2}{n_2 - 1}}$$
And we know from above that:

* the variance of Group A, $s^2_{x_1} = `r var_a`$
* the variance of Group B, $s^2_{x_2} = `r var_b`$
* the N of Group A, $n_1 = `r n_a`$
* and the N of Group B, $n_1 = `r n_b`$

And we can start to fill that in as follows:

$$df = \frac{\left(\frac{`r var_a`}{`r n_a`}+\frac{`r var_b`}{`r n_b`}\right)^2}{\frac{\left(\frac{`r var_a`}{`r n_a`}\right)^2}{`r n_a` - 1}+\frac{\left(\frac{`r var_b`}{`r n_b`}\right)^2}{`r n_b` - 1}}$$

$$df = \frac{\left(\frac{`r var_a`}{`r n_a`}+\frac{`r var_b`}{`r n_b`}\right)^2}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a` - 1}+\frac{\left(\frac{`r var_b`}{`r n_b`}\right)^2}{`r n_b` - 1}}$$
$$df = \frac{\left(\frac{`r var_a`}{`r n_a`}+\frac{`r var_b`}{`r n_b`}\right)^2}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a` - 1}+\frac{\left(`r var_b/n_b`\right)^2}{`r n_b` - 1}}$$

$$df = \frac{\left(\frac{`r var_a`}{`r n_a`}+\frac{`r var_b`}{`r n_b`}\right)^2}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a - 1`}+\frac{\left(`r var_b/n_b`\right)^2}{`r n_b - 1`}}$$
$$df = \frac{\left(`r var_a/n_a`+`r var_b/n_b`\right)^2}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a - 1`}+\frac{\left(`r var_b/n_b`\right)^2}{`r n_b - 1`}}$$

$$df = \frac{`r (var_a/n_a) + (var_b/n_b)`^2}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a - 1`}+\frac{\left(`r var_b/n_b`\right)^2}{`r n_b - 1`}}$$

$$df = \frac{`r ((var_a/n_a)+(var_b/n_b))^2`}{\frac{\left(`r var_a/n_a`\right)^2}{`r n_a - 1`}+\frac{\left(`r var_b/n_b`\right)^2}{`r n_b - 1`}}$$
$$df = \frac{`r ((var_a/n_a)+(var_b/n_b))^2`}{\frac{`r (var_a/n_a)^2`}{`r n_a - 1`}+\frac{`r (var_b/n_b)^2`}{`r n_b - 1`}}$$

$$df = \frac{`r ((var_a/n_a)+(var_b/n_b))^2`}{`r ((var_a/n_a)^2)/(n_a - 1)`+`r ((var_b/n_b)^2)/(n_b - 1)`}$$

$$df = \frac{`r ((var_a/n_a)+(var_b/n_b))^2`}{`r (((var_a/n_a)^2)/(n_a - 1))+ (((var_b/n_b)^2)/(n_b - 1))`}$$

$$df = `r (((var_a/n_a)+(var_b/n_b))^2)/((((var_a/n_a)^2)/(n_a - 1))+ (((var_b/n_b)^2)/(n_b - 1)))`$$